\Large Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada

### 2025-05-09 writing/appendix-a-cleaning-telemetry-data.Rmd
shortened appendix; finished addressing comments

---
title: \Large Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada
subtitle: "Appendix A: Cleaning telemetry data"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'freshwater-biology.csl'
# csl: 'nature.csl'
fontsize: 12pt
# indent: true
header-includes:
    - \renewcommand{\figurename}{Figure S\!\!} # for "Figure Ax.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage{float} # to use fig.pos="H"
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    latex_engine: xelatex # for Delta symbol in table
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
---

```{r setup, include=FALSE}
# set chunk defaults
knitr::opts_chunk$set(echo = TRUE, out.width = '100%',
                      fig.align = 'center', cache = TRUE, warning = FALSE,
                      message = FALSE,
                      root.dir = '~/GitHub/bc-mammals-temperature')
```

@@

\tableofcontents

@@

# Setup

We start by attaching the necessary packages and sourcing some custom functions. The custom functions are all available on GitHub in the `functions` folder of the repository located at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature). The original cleaning script is located at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R). We suggest looking at the script since this appendix does not include the interactive maps that are generated with `plot_adj(..., map = TRUE)`, since they could not be included in the pdf output easily. However, the interacive maps often show linear features (e.g., roads) and other geographic features (e.g., mountain ridges, rivers) that were useful to see during the data cleaning. Therefore, we have left the function calls in the document for completeness, but we have commented them out to prevent them from being executed. Below is an example of a map with some locations for wolf BW008, as generated by running\newline`plot_adj('BW008', max_speed = 3.5, map = TRUE)`.

```{r map-example, echo=FALSE, out.height='40%'}
setwd('writing')
knitr::include_graphics(path = '../figures/BW008-map-example.jpg')
```

@@

@@

@@

```{r final-checks}
# ensure animal IDs are unique
sum(duplicated(d$animal))

# check which species have DOP values
d %>%
  unnest(tel) %>%
  group_by(species) %>%
  summarise(across(c(dop, hdop, pdop), ~ sum(.x, na.rm = TRUE))) %>%
  mutate(has_dop = dop + hdop + pdop > 0) %>%
  arrange(desc(has_dop))

@@

@@

# *Canis lupus*

```{r canis-lupus, fig.pos='!H', fig.dim=c(10, 12)}
# BW008
out <- check_animal('BW008')
plot_adj('BW008', max_speed = 3.5)
# plot_adj('BW008', max_speed = 3.5, map = TRUE)
plot_adj('BW008', max_angle = 178, max_speed = 1) # using a linear feature
plot_adj('BW008', max_angle = 90, max_speed = 2) # realistic movement
plot_adj('BW008', max_angle = 170, max_speed = 0.5) # fine sampling
out[out$angle > 170 & out$speed > 0.5, ] # 15-minute sampling

# BW009
out <- check_animal('BW009')
plot_adj('BW009', max_angle = 179, max_speed = 1) # potential outlier
plot_adj('BW009', max_angle = 179.7, max_speed = 1, n_adj=200) # uses path
plot_adj('BW009', max_angle = 175, max_speed = 1, n_adj = 12) # ok
plot_adj('BW009', max_angle = 170, min_angle = 175, max_speed = 1,
         n_adj = 20) # ok
plot_adj('BW009', max_angle = 170, max_speed = 0.5) # reasonable

@@

# BW012
out <- check_animal('BW012')
plot_adj('BW012', max_speed = 3)
plot_adj('BW012', max_speed = 1, max_angle = 170)
# location is ok
tel <- as.telemetry(d$tel[[which(d$animal == 'BW012')]])
i <- which(out$speed > 2 & out$angle > 170)
plot(tel, error = FALSE, type = 'l', xlim = c(-2600, 0),
     ylim = c(300, 2200), col = 'black')
plot(tel[i + -10:10, ],
     error = FALSE, type = 'l', col = 'blue', lwd = 2, add = TRUE)
plot(tel[i, ],
     error = FALSE, col = 'blue', pch = 19, add = TRUE)
plot(tel, add = TRUE)
rm(tel, i)

# BW013
out <- check_animal('BW013')
plot_adj('BW013', max_speed = 1, max_angle = 170) # uses linear features

@@

# *Cervus canadensis*

```{r cervus-canadensis, fig.pos='!H', fig.dim=c(10, 12)}
#' *NOTE:* assuming movement at ~< 0.25 m/s is realistic, even if angle >
#' 170 degrees since it's too hard to distinguish between movement and gps
#' error at that scale, and the models account for GPS error anyway (which
#' I assume to be 10 m since I don't have calibration data. Mike said most
#' collars have an error ~10 m anyway).

# E002
out <- check_animal('E002')
plot_adj('E002', max_speed = 0.4)
plot_adj('E002', max_angle = 170, max_speed = 0.2)

# E003
out <- check_animal('E003')
plot_adj('E003', max_speed = 0.5)

# E006
out <- check_animal('E006')
plot_adj('E006', max_speed = 0.8) # ok
plot_adj('E006', max_speed = 0.4, max_angle = 170) # outlier
#plot_adj('E006', max_speed = 0.4, max_angle = 170, map = TRUE) # see river
# dt is much larger: collar struggled to receive signal with tree cover
'hours' %#% out[which(out$speed > 0.4 & out$angle > 170) + (-4:4), 'dt']
'hours' %#% c(43282, 7254)
flag_outlier('E006', max_speed = 0.4, max_angle = 170, value = 1)
out <- check_animal('E006')
plot_adj('E006', max_speed = 0.2, max_angle = 170) # ok

# E007
out <- check_animal('E007')
plot_adj('E007', max_speed = 0.2, max_angle = 170) # ok

# E008
out <- check_animal('E008')
plot_adj('E008', max_speed = 0.3) # ok

# E011: last point is outlier (keeping previous points bc many 0 speeds)
out <- check_animal('E011')
plot_adj('E011', max_speed = 0.4, n_adj = 10)
# plot_adj('E011', max_speed = 0.4, n_adj = 10, map = TRUE)
plot_adj('E011', max_speed = 0.4, n_adj = 9, reset_layout = FALSE)
which(out$speed > 0.4) == nrow(out)
i_tel <- which(d$animal == 'E011')
i <- (nrow(d$tel[[i_tel]]) - 7):nrow(d$tel[[i_tel]]) # drop last 8 points
points(location.lat ~ location.long, d$tel[[i_tel]][i, ], col = 'blue',
       cex = 2, pch = 19)
d$tel[[i_tel]][i, 'outlier'] <- 1
points(location.lat ~ location.long, d$tel[[i_tel]][i, ],
       col = as.numeric(d$tel[[i_tel]]$outlier[i]) + 1, cex = 2)
layout(1)

out <- check_animal('E011')
tel <- as.telemetry(d$tel[[i_tel]], mark.rm = TRUE)
plot(tel[nrow(tel) + -10:0, ], error = FALSE, type = 'b',
     pch = as.character(0:9))
plot_adj('E011', max_speed = 0.25)

# E013
out <- check_animal('E013')
plot_adj('E013', max_speed = 0.2, max_angle = 170) # escaped something?

# E015 has multiple GPS malfunctions
out <- check_animal('E015')
plot_adj('E015', max_speed = 1, reset_layout = FALSE)
# can't filter easily because the outliers don't have the max velocity
i <- d %>%
  pull(tel) %>%
  nth(which(d$animal == 'E015')) %>%
  mutate(i = 1:n()) %>%
  slice(which(timestamp ==
                as.POSIXct(out[which(out$speed > 0.6)[2], 't'])) +
          0:1) %>%
  pull(i) %>%
  first() + 0:10
points(location.lat ~ location.long, cex = 2, lwd = 3, col = 'blue',
       filter(d, animal == 'E015')$tel[[1]][i, ])
lines(location.lat ~ location.long,
      filter(d, animal == 'E015')$tel[[1]][i, ], lwd = 6, col = 'red3')
d$tel[[which(d$animal == 'E015')]][i, 'outlier'] <- 1
out <- check_animal('E015')
plot_adj('E015', max_speed = 0.60) # track looks ok now

# E016
out <- check_animal('E016')
plot_adj('E016', max_speed = 0.2, max_angle = 170) # one outlier
plot_adj('E016', max_speed = 0.2, max_angle = 179) # use higher upper bound
flag_outlier('E016', max_speed = 0.2, max_angle = 179, value = 1)
out <- check_animal('E016')
plot_adj('E016', max_speed = 0.2, max_angle = 170) # ok

# E017
out <- check_animal('E017')
plot_adj('E017', max_speed = 0.6) # ok
plot_adj('E017', max_angle = 170, max_speed = 0.2)

# E019
out <- check_animal('E019') # many high speeds at angle > 170
plot_adj('E019', max_speed = 0.65, max_angle = 135) # some outliers
plot_adj('E019', max_speed = 0.65, max_angle = 160, reset_layout = FALSE)
i <- which(out$speed > 0.65 & out$angle > 160)[c(1, 3)] # some outliers
points(location.lat ~ location.long,
       d$tel[[which(d$animal == 'E019')]][i, ], cex = 2, col = 'blue')
d$tel[[which(d$animal == 'E019')]][i, 'outlier'] <- 1

out <- check_animal('E019') # check other high speeds with angle > 160
plot_adj('E019', max_speed = 0.4, max_angle = 135, reset_layout = FALSE)
tel <- d$tel[[which(d$animal == 'E019')]]
i <- which(out$speed > 0.4 & out$angle > 135)
points(location.lat ~ location.long, cex = 1.5, col = 'black',
       tel[! tel$outlier, ][i, ],
       pch = as.character(0:(length(i) - 1)))
'hour' %#% out[i, 'dt'] # consistent sampling

# 0 is likely an error (erring on the side of caution)
# 1 is consistent with other movement
# 2 is likely an error
# 3 is consistent with other movement
# 4 is likely an error
# 5 is realistic
# 6 is likely an error
# 7 is realistic
# 8 is likely an error
# 9 is likely an error
i <- i[1 + c(0, 2, 4, 6, 8, 9)]
i_id <- which(d$animal == 'E019')
d$tel[[i_id]][! d$tel[[i_id]]$outlier, ][i, 'outlier'] <- 1
# ensure correct locations were removed
out <- check_animal('E019') # check other high speeds with angle > 160
plot_adj('E019', max_speed = 0.4, max_angle = 135, reset_layout = FALSE)

# E023
out <- check_animal('E023')
plot_adj('E023', max_speed = 1) # ok
plot_adj('E023', max_speed = 0.6, min_speed = 1) # ok
plot_adj('E023', max_speed = 0.3, max_angle = 170) # ok

# E024
out <- check_animal('E024')
plot_adj('E024', max_speed = 0.4) # ok
plot_adj('E024', max_speed = 0.2, max_angle = 170) # ok

# E025
out <- check_animal('E025')
plot_adj('E025', max_speed = 0.3, max_angle = 170) # ok

# E026
#' this elk does not have an unusual amount of data
hist(map_int(filter(d, species == 'Cervus canadensis')$tel, nrow))
abline(v = nrow(d$tel[[which(d$animal == 'E026')]]), lwd = 2)
# d$tel[[which(d$animal == 'E026')]] %>%
#   filter(! outlier) %>%
#   sf::st_as_sf(coords = c('location.long', 'location.lat'),
#                crs = 4326) %>%
#   mapview(map.types = 'Esri.WorldImagery', col.regions = 'darkorange',
#           col = 'black', lwd = 2)
out <- check_animal('E026') # many sharp turns at high speeds
plot_adj('E026', max_speed = 0.6, max_angle = 135)
#plot_adj('E026', max_speed = 0.6, max_angle = 135, map = TRUE)# look north
flag_outlier('E026', max_speed = 0.6, max_angle = 135, value = 1)

out <- check_animal('E026')
plot_adj('E026', max_speed = 0.4, max_angle = 170) # NW one is ok
# plot_adj('E026', max_speed = 0.4, max_angle = 170, map = TRUE)
plot_adj('E026', max_speed = 0.42, max_angle = 170) # W one may also be ok
plot_adj('E026', max_speed = 0.42, max_angle = 175) # W one may also be ok
flag_outlier('E026', max_speed = 0.42, max_angle = 175, value = 1)

out <- check_animal('E026')
plot_adj('E026', max_speed = 0.35, max_angle = 170) # NW, middle, & NE ok?
plot_adj('E026', max_speed = 0.35, min_speed = 0.4, max_angle = 170,
         reset_layout = FALSE)
i <- with(out, which(speed > 0.35 & speed < 0.4 & angle > 160))[c(3:4)]
points(location.lat ~ location.long,
       filter(d$tel[[which(d$animal == 'E026')]], ! outlier)[i, ],
       cex = 2, col = 'blue', lwd = 2)
#' values in `i` assume outliers have been removed
i_tel <- which(d$animal == 'E026')
d$tel[[i_tel]][d$tel[[i_tel]]$outlier == 0, ][i, 'outlier'] <- 1

out <- check_animal('E026')
plot_adj('E026', max_speed = 0.30, max_angle = 160) # likely ok

#' using an error-informed, continuous-time movement model substantially
#' reduces the issues caused by GPS error. Additionally, there are so many
#' elk data (for each individual and the number of elk, too) that a few
#' values from a single individual will have little to no effect.

# E027
out <- check_animal('E027')
plot_adj('E027', max_speed = 0.3, max_angle = 170, n_adj = 4) # not all
plot_adj('E027', max_speed = 0.3, max_angle = 175, n_adj = 4,
         reset_layout = FALSE)
i <- with(out, which(speed > 0.3 & angle > 175))[c(2, 3, 4, 5)]
points(location.lat ~ location.long,
       d$tel[[which(d$animal == 'E027')]][i, ],
       cex = 2, col = 'blue', lwd = 2)
'hours' %#% out$dt[i]
d$tel[[which(d$animal == 'E027')]][i, 'outlier'] <- 2 # not fully sure

# E030
out <- check_animal('E030')
plot_adj('E030', max_speed = 0.45) # ok
plot_adj('E030', max_speed = 0.3, max_angle = 135) # generally ok
plot_adj('E030', max_speed = 0.3, max_angle = 170) # possibly problematic
flag_outlier('E030', max_speed = 0.3, max_angle = 170, value = 1)

out <- check_animal('E030')
plot_adj('E030', max_speed = 0.2, max_angle = 170) # ok

# E032
out <- check_animal('E032')
plot_adj('E030', max_speed = 0.3, max_angle = 170) # ok

# E033
out <- check_animal(id = 'E033')
plot_adj('E033', max_speed = 0.6)

# E036
out <- check_animal(id = 'E036')
# the next point is along the same line 
plot_adj('E036', max_speed = 0.7, reset_layout = FALSE)
points(location.lat ~ location.long, pch = 19,
       filter(d, animal == 'E036')$tel[[1]][0:1 + which(out$speed > 0.7),])
plot_adj('E036', max_speed = 0.6, reset_layout = FALSE) # ok

# E037
out <- check_animal(id = 'E037')
plot_adj(id = 'E037', max_speed = 0.3, max_angle = 170) # ok

# E040
out <- check_animal(id = 'E040')
plot_adj('E040', max_speed = 0.35, max_angle = 160) # ok

# E042: range shift with highly stationary behavior
out <- check_animal(id = 'E042')
tel <- as.telemetry(d$tel[[which(d$animal == 'E042')]])
range(tel$timestamp)
plot(tel)
# plot_adj(id = 'E042', many = TRUE, map = TRUE, max_speed = 0)
ctmm.guess(tel)

# E049
out <- check_animal(id = 'E049')
plot_adj('E049', max_speed = 0.4, max_angle = 170) # ok
plot_adj('E049', max_speed = 0.3, min_speed = 0.4, max_angle = 170) # ok

# E067
out <- check_animal(id = 'E067')
plot_adj('E067', max_speed = 0.6, max_angle = 170) # ok

# E068
out <- check_animal(id = 'E068')
plot_adj('E068', max_speed = 0.65, reset_layout = FALSE)
out[out$speed > 0.65, ] # 1st 2 points are right after capture (rownames)
points(location.lat ~ location.long,
       filter(d, animal == 'E068')$tel[[1]][1:2, ], col = 'blue', cex = 2)
d[d$animal == 'E068', ]$tel[[1]]$outlier[1:2] <- 1
out <- check_animal(id = 'E068')
plot_adj('E068', max_speed = 0.4, max_angle = 170)
# plot_adj('E068', max_speed = 0.4, max_angle = 170, map = TRUE)
flag_outlier('E068', max_speed = 0.4, max_angle = 170, value = 1)
out <- check_animal(id = 'E068')
plot_adj('E068', max_speed = 0.25, max_angle = 170) # ok

# E070
out <- check_animal('E070')
plot_adj('E070', max_speed = 0.4, max_angle = 135) # ok

# E074
out <- check_animal(id = 'E074')
plot_adj('E074', max_speed = 0.4, max_angle = 170) # ok

# E075
out <- check_animal(id = 'E075')
plot_adj('E075', max_speed = 0.3, max_angle = 170) # ok

# E079
out <- check_animal(id = 'E079')
plot_adj('E079', max_speed = 0.6) # ok

# E080
out <- check_animal(id = 'E080')
plot_adj('E080', max_speed = 0.8) # ok
plot_adj('E080', max_speed = 0.5, max_angle = 170, reset_layout = FALSE)
i_tel <- which(d$animal == 'E080')
i <- which(out$speed > 0.5 & out$angle > 170)[1:2]
points(location.lat ~ location.long,
       filter(d$tel[[i_tel]], ! outlier)[i, ], col = 'blue',
       cex = 2)
d$tel[[i_tel]][! d$tel[[i_tel]]$outlier, ]$outlier[i] <- 1
out <- check_animal(id = 'E080')
plot_adj('E080', max_speed = 0.5, max_angle = 170) # ok

# E081
out <- check_animal(id = 'E081')
plot_adj('E081', max_speed = 0.2, max_angle = 170) # ok

# E082
out <- check_animal(id = 'E082')
plot_adj('E082', max_speed = 0.2, max_angle = 170) # ok

# E083
out <- check_animal(id = 'E083')
plot_adj('E083', max_speed = 0.3, max_angle = 135) # ok

# E084 is not worth keeping
out <- check_animal(id = 'E084')
plot(location.lat ~ location.long, filter(d, animal == 'E084')$tel[[1]],
     type = 'b', pch = 19, col = '#00000010')
nrow(filter(d, animal == 'E084')$tel[[1]])
range(filter(d, animal == 'E084')$tel[[1]]$timestamp)
d <- filter(d, animal != 'E084')

# E085
out <- check_animal(id = 'E085')
plot_adj('E085', max_speed = 1) # definitely outliers (> 20 km!)
flag_outlier(id = 'E085', max_speed = 1, value = 1)

out <- check_animal(id = 'E085')
plot_adj('E085', max_speed = 0.6) # maybe ok
with(out[which(out$speed > 0.6) + -10:10, ], # likely outlier
     plot('hours' %#% dt, dt * speed / 1e3,
          col = as.numeric(-10:10 == 0) + 1,
          pch = 19, ylab = 'SLD (dt * min speed, km)'))
flag_outlier('E085', max_speed = 0.6, value = 1)

out <- check_animal(id = 'E085')
plot_adj('E085', max_speed = 0.2, max_angle = 135) # ok

# E096
out <- check_animal(id = 'E096')
plot_adj('E096', max_speed = 0.4, max_angle = 170) # ok

# E100
out <- check_animal(id = 'E100')
plot_adj('E100', max_speed = 0.3, max_angle = 170) # ok

# E101
out <- check_animal(id = 'E101')
plot_adj('E101', max_speed = 0.4, max_angle = 170) # ok

# E106
d$tel[[which(d$animal == 'E106')]] %>%
  group_by(gps.fix.type.raw) %>%
  summarise(hdop = mean(dop, na.rm = TRUE))
d$tel[[which(d$animal == 'E106')]] %>%
  group_by(gps.fix.type.raw) %>%
  summarise(hdop = mean(! is.na(dop), na.rm = TRUE))
out <- check_animal(id = 'E106')
plot_adj('E106', max_speed = 0.4, max_angle = 170) # large DOP values; ok
d$tel[[which(d$animal == 'E106')]] %>%
  as.telemetry() %>%
  data.frame() %>%
  slice(which(out$speed > 0.4 & out$angle > 170)) %>%
  pull(HDOP)
plot_adj('E106', max_speed = 0.3, max_angle = 170) # ok
plot_adj('E106', max_speed = 0.2, max_angle = 170) # ok

# E115
# collar is very error-prone in SW forest
out <- check_animal(id = 'E115')
# plot_adj('E115', many = TRUE, map = TRUE)
plot_adj('E115', max_speed = 0.4, max_angle = 170) # improbable
plot_adj('E115', max_speed = 0.4, max_angle = 170, n_adj = 50)#but possible
plot_adj('E115', max_speed = 0.4, max_angle = 175) # unlikely
slice(d$tel[[which(d$animal == 'E115')]],
      which(out$speed > 0.4 & out$angle > 175))$dop # low DOP
flag_outlier(id = 'E115', max_speed = 0.4, max_angle = 175, value = 2)

out <- check_animal('E115')
plot_adj('E115', max_speed = 0.3, max_angle = 170) # unlikely
flag_outlier('E115', max_speed = 0.3, max_angle = 170, value = 2)

out <- check_animal('E115')
plot_adj('E115', max_speed = 0.1, max_angle = 170, many = TRUE) # ok

# E116
out <- check_animal(id = 'E116')
plot_adj('E116', max_speed = 0.4, max_angle = 170)
flag_outlier(id = 'E116', max_speed = 0.4, max_angle = 170, value = 1)

out <- check_animal('E116')
plot_adj('E116', max_speed = 0.8)
flag_outlier(id = 'E116', max_speed = 0.8, value = 1)

out <- check_animal('E116')
plot_adj('E116', max_speed = 0.3, max_angle = 170) # improbable
plot_adj('E116', max_speed = 0.3, max_angle = 175) # but not impossible

# E117 has similar error patterns to E115
out <- check_animal(id = 'E117')
plot_adj('E117', max_speed = 0.7) # three outliers
# plot_adj('E117', max_speed = 0.7, map = TRUE)
flag_outlier(id = 'E117', max_speed = 0.7, value = 1)

out <- check_animal('E117')
#plot_adj('E117', many = TRUE, map = TRUE) # more errors when more forested
plot_adj('E117', max_speed = 0.4, max_angle = 170)

# plot_adj('E117', max_speed = 0.4, max_angle = 170, map = TRUE)
flag_outlier(id = 'E117', max_speed = 0.4, max_angle = 170, value = 1)

out <- check_animal('E117')
plot_adj('E117', max_speed = 0.3, max_angle = 170) # maybe not all
plot_adj('E117', max_speed = 0.3, max_angle = 175) # likely all
#plot_adj('E117', max_speed = 0.3, max_angle = 175, map = TRUE) # in forest
flag_outlier(id = 'E117', max_speed = 0.3, max_angle = 175, value = 1)

out <- check_animal('E117')
# still an excessive amount of sharp turns with high speeds (compare E118)
# dropping some of the data is better than dropping the whole animal
flag_outlier('E117', max_speed = 0.1, max_angle = 130, value = 2)
out <- check_animal('E117')

# E118 (compare with E117 above)
check_animal('E118', return_out = FALSE)

# E122
out <- check_animal(id = 'E122')
plot_adj('E122', max_speed = 0.45) # ok

# E125
out <- check_animal(id = 'E125')
plot_adj('E125', max_speed = 0.4, max_angle = 150)
flag_outlier('E125', max_speed = 0.4, max_angle = 150, value = 1)
out <- check_animal(id = 'E125') # ok

# E129
out <- check_animal(id = 'E129')
plot_adj('E129', max_speed = 0.6) # ok

# E131
out <- check_animal(id = 'E131')
plot_adj('E131', max_speed = 0.3, max_angle = 170) # N one is ok
plot_adj('E131', max_speed = 0.45, max_angle = 170) # outlier
flag_outlier('E131', max_speed = 0.45, max_angle = 170, value = 1)

out <- check_animal(id = 'E131') # ok

# E132
#' some high minimum speeds at long sampling intervals for this individual
out <- check_animal(id = 'E132', cap_dt = FALSE)
layout(matrix(c(0, 0, 0, 1), ncol = 2)) #
abline(v = 20, col = 'red')
abline(h = 0.08, col = 'red')
i <- which(out$dt > 20 %#% 'hours' & out$speed > 0.08)
out[i, ]
# small range shift at the daily scale within home range
plot_adj('E132', max_speed = 0.08, max_dt = 20 %#% 'hours')

# look at points near values with excessively high speeds
i_adj <- sapply(i, function(.i) .i + -10:10) %>%
  as.data.frame() %>% # convert matrix to a data frame
  pivot_longer(! c()) # use long format for a column of all values
tel <- as.telemetry(d$tel[[which(d$animal == 'E132')]])
adj <- left_join(slice(data.frame(out), i_adj$value),
                 slice(data.frame(tel), unique(i_adj$value)),
                 by = c('t')) %>%
  bind_cols(i_adj, .) %>%
  mutate(date = as.POSIXct(t),
         high_speed = speed > 0.08)
layout(matrix(1:4, ncol = 2))
for(g in seq(unique(i_adj$name))) {
  g_name <- unique(i_adj$name)[g]
  plot(y ~ x, filter(adj, name == g_name), type = 'l')
  points(y ~ x, data.frame(tel[i, ][g, ]), col = 'blue', pch = 19)
  rm(g, g_name)
} # ok
layout(1)

# large dt with minor "range shifts" ==> ok
plot('days' %#% dt ~ as.POSIXct(t), out[i_adj$value, ])
points('days' %#% dt ~ as.POSIXct(t), out[i, ], col = 'blue', pch = 19)

rm(i, i_adj, tel, adj)

# E134
out <- check_animal(id = 'E134')
plot_adj('E134', max_speed = 0.5, max_angle = 170) # ok

# E138
out <- check_animal(id = 'E138')
plot_adj('E138', max_speed = 0.8) # ok
plot_adj('E138', max_speed = 0.3, max_angle = 170) # ok

# E139
out <- check_animal(id = 'E139')
plot_adj('E139', max_speed = 0.3, max_angle = 170) # ok

# E140
out <- check_animal(id = 'E140')
plot_adj('E140', max_speed = 0.6)
flag_outlier('E140', max_speed = 0.6, value = 1)

out <- check_animal(id = 'E140') # ok

# E142
out <- check_animal(id = 'E142')
flag_outlier('E142', max_speed = 10, value = 1) # definitely an outlier

out <- check_animal(id = 'E142')
plot_adj('E142', max_speed = 0.8) # ok

# E143
out <- check_animal(id = 'E143')
plot_adj('E143', max_speed = 0.2, max_angle = 170) # ok

# E144
out <- check_animal(id = 'E144')
plot_adj('E144', max_speed = 0.3, max_angle = 170) # all ok

>>>>

---
title: \Large Predicting how climate change will affect how and where terrestrial mammals will move in British Columbia, Canada
subtitle: "Appendix A: Cleaning telemetry data"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: [braes, biol]
  - name: Chris H. Fleming
    institute: [ucf, smith]
  - name: Siobhan Darlington
    institute: [braes, biol]
  - name: Adam T. Ford
    institute: [braes, biol]
  - name: Karen E. Hodges
    institute: [braes, biol]
  - name: Kirk Safford
    institute: bcp
  - name: Robert Serrouya
    institute: [braes, biol, biodiv]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: [braes, biol, cmps]
institute:
  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
  - bcp: BC Parks
  - biodiv: Wildlife Science Centre, Biodiversity Pathways, University of British Columbia Okanagan, Revelstoke, British Columbia, Canada.
  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'bc-mammals-temperature.bib'
csl: 'movement-ecology.csl'
# csl: 'nature.csl'
fontsize: 12pt
# indent: true
header-includes:
    - \renewcommand{\figurename}{Figure S\!\!} # for "Figure Ax.
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
    - \usepackage{hanging} # for hanging indents in references
    - \usepackage{float} # to use fig.pos="H"
subparagraph: true # needed for \usepackage[small]{titlesec}
urlcolor: cyan
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: true
---

```{r setup, include=FALSE}
# set chunk defaults
knitr::opts_chunk$set(echo = TRUE, out.width = '100%',
                      fig.align = 'center', cache = TRUE, warning = FALSE,
                      message = FALSE,
                      root.dir = 'H:/GitHub/bc-mammals-temperature')
```

@@

In this appendix, we illustrate the cleaning procedure for the telemetry data. For the sake of brevity, we only show the code and outputs for *Canis lupus*; the full script is available at  [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature/blob/main/analysis/02-remove-outliers.R). We suggest looking at the script since this appendix does not include the interactive maps that are generated by `plot_adj(..., map = TRUE)`. The interactive maps often show linear features (e.g., roads) and other geographic features (e.g., mountain ridges, rivers) that were useful to see during the data cleaning. Therefore, we have left the function calls in the document for completeness, but we have commented them out to prevent them from being executed. Below is an example of a map with some locations for wolf BW008, as generated by running\newline`plot_adj('BW008', max_speed = 3.5, map = TRUE)`.

```{r map-example, echo=FALSE, out.height='40%'}
knitr::include_graphics(path = 'H:/GitHub/bc-mammals-temperature/figures/BW008-map-example.jpg',
                        rel_path = FALSE)
```

We start by attaching the necessary packages and sourcing some custom functions. The custom functions are all available on GitHub in the `functions` folder of the repository located at [https://github.com/QuantitativeEcologyLab/bc-mammals-temperature](https://github.com/QuantitativeEcologyLab/bc-mammals-temperature).

@@

\small

@@

\normalsize

@@

\small

@@

\small

```{r final-checks}
# ensure animal IDs are unique
sum(duplicated(d$animal))
```

```{r final-checks2}
# check which species have DOP values
d %>%
  unnest(tel) %>%
  group_by(species) %>%
  summarise(across(c(dop, hdop, pdop), ~ sum(.x, na.rm = TRUE))) %>%
  mutate(has_dop = dop + hdop + pdop > 0) %>%
  arrange(desc(has_dop))

@@

\normalsize

Now that we have organized the full telemetry dataset, we can start cleaning the data.

@@

\small

```{r canis-lupus, fig.pos='!H', fig.dim=c(10, 12)}
# BW008
out <- check_animal('BW008')
plot_adj('BW008', max_speed = 3.5)
# plot_adj('BW008', max_speed = 3.5, map = TRUE)
plot_adj('BW008', max_angle = 178, max_speed = 1) # using a linear feature
plot_adj('BW008', max_angle = 90, max_speed = 2) # realistic movement
plot_adj('BW008', max_angle = 170, max_speed = 0.5) # fine sampling
```

```{r canis-lupus-2, fig.pos='!H', fig.dim=c(10, 12)}
out[out$angle > 170 & out$speed > 0.5, ] # 15-minute sampling
```

```{r canis-lupus-3, fig.pos='!H', fig.dim=c(10, 12)}
# BW009
out <- check_animal('BW009')
plot_adj('BW009', max_angle = 179, max_speed = 1) # potential outlier
plot_adj('BW009', max_angle = 179.7, max_speed = 1, n_adj=200) # uses path
plot_adj('BW009', max_angle = 175, max_speed = 1, n_adj = 12) # ok
plot_adj('BW009', max_angle = 170, min_angle = 175, max_speed = 1,
         n_adj = 20) # ok
plot_adj('BW009', max_angle = 170, max_speed = 0.5) # reasonable

@@

# BW012
out <- check_animal('BW012')
plot_adj('BW012', max_speed = 3)
plot_adj('BW012', max_speed = 1, max_angle = 170)
```

```{r canis-lupus-4, fig.pos='!H', fig.dim=c(5, 6)}
# location is ok
tel <- as.telemetry(d$tel[[which(d$animal == 'BW012')]])
i <- which(out$speed > 2 & out$angle > 170)
plot(tel, error = FALSE, type = 'l', xlim = c(-2600, 0),
     ylim = c(300, 2200), col = 'black')
plot(tel[i + -10:10, ],
     error = FALSE, type = 'l', col = 'blue', lwd = 2, add = TRUE)
plot(tel[i, ],
     error = FALSE, col = 'blue', pch = 19, add = TRUE)
plot(tel, add = TRUE)
rm(tel, i)
```

```{r canis-lupus-5, fig.pos='!H', fig.dim=c(10, 12)}
# BW013
out <- check_animal('BW013')
plot_adj('BW013', max_speed = 1, max_angle = 170) # uses linear features

@@
